---
title: "Class 15"
author: "Divyanshu Kawankar"
date: "11/20/2021"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

#Important countData and colData

```{r}
#Count Data and read Metadata files

library(BiocManager)

library(DESeq2)

counts <- read.csv("airway_scaledcounts.csv", row.names=1)

metadata <-  read.csv("airway_metadata.csv")

#Take a look at each

head(counts)

head(metadata)

dim(counts)
```
> Q1. How many genes are in this dataset?
 - 38694
 
> Q2. How many ‘control’ cell lines do we have?
 - 4
 
#Toy Differential Gene Expression

```{r}
#Find the sample id for the labeled control
control <- metadata[metadata[,"dex"]=="control",]

control.counts <- counts[ ,control$id]

control.mean <- rowSums( control.counts )/4 

head(control.mean)

#also the sample id for the treated
treated <- metadata[metadata[,"dex"]=="treated",]

treated.counts <- counts[ ,control$id]

treated.mean <- rowSums( treated.counts )/4 

head(treated.mean)
```

 > Q3. How would you make the above code in either approach more robust?
 - answer is below

```{r}
#this is an example of how the code would be more robust. let "smple" be represented whatever variable you are using. I used "control" for my example because that's what we used in the lab.Replace the number in the code with however many variables you are analyzing. 
sampleid_sample<-function(ex)  {
  
library (dplyr)
  
control <- metadata[metadata[,"dex"]==ex,]

control.counts <- counts[ ,control$id]

control.mean <- rowSums( control.counts )/4 

head(control.mean)
}

#plug in "control" to see it in action
sampleid_sample("control")
```
```{r}
#combine data frame

meancounts <- data.frame(control.mean, treated.mean)

colSums(meancounts)
```

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)
  - answer is below

```{r}
#use the same equation from above but plug in "treated"
sampleid_sample("treated")
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

 > Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 
 - geom_point
  
```{r}
library(ggplot2)

geom_point(plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated"))
```

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?
  - argument is "log"

```{r}
plot(meancounts[,1],meancounts[,2], log="xy", xlab="Control", ylab="Treated")
```

```{r}
#calculate log2foldchange
meancounts$log2fc <-log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
#inspect resuls and intermediate step
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

 > Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?
 - The arr.ind argument allows for us to see the results with only true values. We focus only on the row answer. The unique() function will make sure that we don't see any repeats
 
> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 
 - 250
 
> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 
 - 367
 
> Q10. Do you trust these results? Why or why not?
 - We see differences but there is no way to know if it a significant difference or if they are too similar to be considered different.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)

length(which(up.ind, arr.ind=TRUE))

length(which(down.ind, arr.ind=T))
```

#DeSeq2 Analysis
```{r}
library(DESeq2)
citation("DESeq2")
```
 
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

```{r}
dds <- DESeq(dds)

res <- results(dds)

res

summary(res)
```

```{r}
#set alpha value to 0.05

res05 <- results(dds, alpha=0.05)

summary(res05)
```
 
#Adding Annotated Data

```{r}

library("AnnotationDbi")

library("org.Hs.eg.db")

columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     
 keys=row.names(res), 
                     
 keytype="ENSEMBL",
                     
 column="SYMBOL",
                     
 multiVals="first")

head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.
 - Refer to the code below 
 
```{r}
res$entrez <- mapIds(org.Hs.eg.db,
          keys=row.names(res),
          column="ENTREZID",
          keytype="ENSEMBL",
          multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
          keys=row.names(res),
          column="UNIPROT",
          keytype="ENSEMBL",
          multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
          keys=row.names(res),
          column="GENENAME",
          keytype="ENSEMBL",
          multiVals="first")

head(res)
```

```{r}
#view results by adcustied c value and save it.
ord <- order( res$padj )

head(res[ord,])

write.csv(res[ord,], "deseq_results.csv")
```

#Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

# Examine the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 2)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
# Look at the first three down (less) pathways
head(keggres$less, 3)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa05310", kegg.native=FALSE)
```

> Q12. Can you do the same procedure as above to plot the pathview figures for the top 2 down-reguled pathways?
 - Yes. Especially if you know the codes for the pathways. 